\chapter*{Acknowledgements}

\noindent Thanks to my supervisors, Professor Janaina Mourao-Miranda and Professor John Shawe-Taylor, for their contributions. I am very grateful to the EPSRC UCL Centre for Doctoral Training (CDT) in Intelligent, Integrated Imaging in Healthcare (i4Health) and NIHR UCLH Biomedical Research Centre for funding this research. Thanks to G-Research for funding my trip to NeurIPS 2023 to present my work.

\bigskip

\noindent To my incredible friends and coauthors, Ana and Lennie, I couldn't have done this without you. Ana, you kept our paper alive when I had given up on it (as well as the PhD). Lennie, your brutal honesty about the work being rubbish made it infinitely better. Florence, I was honored to be included on Fusili and for the hard lessons you taught me in marketing.

\bigskip

\noindent To members of the Machine Learning for Neuroimaging group at UCL, the Centre for Medical Imaging Computing, and all of the friends from 90 High Holborn. Cemre, I wish we had more opportunities to collaborate, but I am grateful for your words of advice before you left. Agoston, I was inspired by your immense knowledge of the field. Rick, you were everpresent in the office whenever the pandemic allowed, and I appreciated your offers to get lunch together. To the Mojo Dojo Casa House, thank you for making my NeurIPS 2023 experience unforgettable.

\bigskip

\noindent To the boat clubs of University College London, University of London, and Vesta Rowing Club who have provided an outlet that gave me a sense of progress, purpose, and community, even when academia had me down. Thanks also to all of my friends who have listened to me complain about my PhD. And thanks to the birds outside the window for the entertainment you provided during breaks from the desk while working from home.

\bigskip

\noindent Thanks to my mum and dad. Obviously this has been a bit of a rollercoaster, but I am grateful to you for supporting me through it. I will be proud to (just about) be the second Dr Chapman in the family.

\bigskip

\noindent And finally, with lots of love to Rebecca. On about our third date, you came to visit Leamington Spa to scout out PhDs. You have experienced every good and bad moment. You have been proud of me. We got through this together and we will get through the next thing together too.