%	Self-plagiarism declaration form template for these typeset in LaTeX
%	Prepared by David Sheard 2022 and made available free of copyright

%	If you use the results of your own published, accepted or submitted data (text or figures) in your final 
%	doctoral thesis, you have to give a clear indication of the previous work, stating the exact source of the 
%	previous material, irrespective of whether copyright is owned by you or by a publisher. This indication 
%	should take the form of 
%		a) an appropriate citation of the original source in the relevant Chapter; and 
%		b) completion of the UCL Research Paper Declaration form---this should be embedded after the 
%		Acknowledgements page in the thesis.

%	For mor information consult the following links:
%	\url{https://www.grad.ucl.ac.uk/essinfo/guidance-on-selfplagiarism/?utm_source=Students\%27+Union+UCL\&utm_campaign=2ed9e73ab7-\&utm_medium=email\&utm_term=0_fe8c0cbcf2-2ed9e73ab7-209240456\&mc_cid=2ed9e73ab7\&mc_eid=0496c22bfc}
%	\url{https://www.grad.ucl.ac.uk/essinfo/guidance-on-selfplagiarism/Declaration-form_published-work-in-thesis.docx}

%	I recommend using this template by simply copying everything between \begin{document} and \end{document} 
%	into your thesis after the acknowledgements section. By changing the preamble appropriately this template 
%	may ablso be altered to work using \input{...} or the subfiles package.

% \documentclass[12pt, twoside]{article}

% \usepackage[a4paper,inner=40mm,outer=20mm, top=30mm, bottom=30mm]{geometry}
% \usepackage{amssymb}
% \usepackage{array}
% \usepackage{setspace}
% 	\setstretch{1.5}

% \begin{document}
	
\section*{UCL Research Paper Declaration Form: referencing the doctoral candidate’s own published work(s)}
	
% Uncomment the following line if you would like to add this declaration to your table of contents
%	\addcontentsline{toc}{section}{UCL Research Paper Declaration Form}

%	Please use this form to declare if parts of your thesis are already available in another format, 
%	e.g. if data, text, or figures:
%	•	have been uploaded to a preprint server 
%	•	are in submission to a peer-reviewed publication 
%	•	have been published in a peer-reviewed publication, e.g. journal, textbook.  

%	This form should be completed as many times as necessary. For instance, if a student had seven 
%	thesis chapters, two of which having material which had been published, they would complete this form twice. 

\begin{enumerate}\itemsep0em
%		
	\item \textbf{1.	For a research manuscript that has already been published} (if not yet published, please skip to section 2)\textbf{:}
%
	\begin{enumerate}\itemsep0em
%			
		\item \textbf{What is the title of the manuscript?}
		Unconstrained Stochastic CCA: Unifying Multiview and Self-Supervised Learning
		\item \textbf{Please include a link to or doi for the work:}
		https://iclr.cc/virtual/2024/poster/18714
		\item \textbf{Where was the work published?}
		Proceedings of the International Conference on Learning Representations
		\item \textbf{Who published the work?}
		International Conference on Learning Representations
		\item \textbf{When was the work published?}
		May 2024
		\item \textbf{List the manuscript's authors in the order they appear on the publication:}
		James Chapman and Lennie Wells and Ana Lawry Aguila
		\item \textbf{Was the work peer reviewd?}
		Yes
		\item \textbf{Have you retained the copyright?}
		Yes
		\item \textbf{Was an earlier form of the manuscript uploaded to a preprint server (e.g. medRxiv)? If ‘Yes’, please give a link or doi} 
		https://arxiv.org/abs/2310.01012
		\\
		If ‘No’, please seek permission from the relevant publisher and check the box next to the below statement:
%			
		\begin{itemize}\itemsep0em
			% To check this box, replace \Box with \boxtimes
			\item[$\boxtimes$] {\itshape I acknowledge permission of the publisher named under 1d to include in this thesis portions of the publication named as included in 1c.}
		\end{itemize}
%		
	\end{enumerate}
%	
	\item \textbf{For a research manuscript prepared for publication but that has not yet been published} (if already published, please skip to section 3)\textbf{:}
%		
	\begin{enumerate}\itemsep0em
		%			
		\item \textbf{What is the current title of the manuscript?}
		% Unconstrained Stochastic CCA: Unifying Multiview and Self-Supervised Learning
		\item \textbf{Has the manuscript been uploaded to a preprint server 'e.g. medRxiv'? 
		\\
		If 'Yes', please please give a link or doi:}
		% Answer here:
		\item \textbf{Where is the work intended to be published?}
		% Answer here: e.g. journal name
		\item \textbf{List the manuscript's authors in the intended authorship order:}
		% Answer here
		\item \textbf{Stage of publication:}
		% answer here: e.g. in submission
	%		
	\end{enumerate}
	
	\item \textbf{For multi-authored work, please give a statement of contribution covering all authors} (if single-author, please skip to section 4)\textbf{:}
	% Answer here
	\item \textbf{In which chapter(s) of your thesis can this material be found?}
	5,6
	%
\end{enumerate}

	\textbf{e-Signatures confirming that the information above is accurate}
	(this form should be co-signed by the supervisor/ senior author unless this is not appropriate, e.g. if the paper was a single-author work)\textbf{:}\\
	\textbf{James Chapman}\\ 
	\textbf{Candidate: James Chapman}\\
	\textbf{Date: 17/04/2024}\\
	% this form should be co-signed by the supervisor/ senior author unless this is not appropriate, e.g. if the paper was a single-author work):
	\textbf{}\\
	\textbf{Supervisor/Senior Author signature} (where appropriate)\textbf{:}\\
	\textbf{Date:}
	%		


% \end{document}
